## Practical 1

1. Reproduce the timing results in chapter 2 using the `mark()`
  function from the `bench` package. Remember to load the package using
    ```{r,echo=TRUE}
    library("bench")
    ```
2. **Case study** In this example, we are going to investigate loading a
  large data frame. First, we'll generate a large matrix of random numbers and
  save it as a csv file:^[If setting `N=1e6` is too large for your
    machine, reduce it at bit. For example, `N=50,000`.]
    ```{r,practical1, cache=TRUE, echo = TRUE}
    N = 1e5
    m = as.data.frame(matrix(runif(N), ncol = 1000))
    write.csv(m, file = "example.csv", row.names = FALSE)
    ```
We can read the file the back in again using `read.csv`
    ```{r,cache=TRUE, echo = TRUE}
    dd = read.csv("example.csv")
    ```
To get a baseline result, time the `read.csv` function call above.
    ```{r,cache=TRUE, echo = TRUE}
    system.time(read.csv("example.csv"))
    ```
\noindent We will now look ways of speeding up this step.
    + Explicitly define the classes of each column using `colClasses` in
    `read.csv`, for example, if we have 1000 columns that all have data
    type numeric, then:
        ```{r,cache=TRUE,results='hide', tidy=FALSE, echo = TRUE}
        read.csv(file = "example.csv",
                 colClasses = rep("numeric", 1000))
        ```
    + Use the `saveRDS` and `readRDS` functions:
        ```{r,cache=TRUE, results="hide", echo = TRUE}
        saveRDS(m, file = "example.RData")
        readRDS(file = "example.RData")
        ```
    + Install the `readr` package via
        ```{r,eval=FALSE, echo = TRUE}
        install.packages("readr")
        ```
        Then load the package in the usual way
        ```{r, echo = TRUE, eval= FALSE}
        library("readr")
        ```
        This package contains the function `read_csv`; a replacement function for `read.cvs`. Is this new function much better than the default?

  Which of the above give the biggest speed-ups? Are there any downsides to using these   techniques? Do your results depend on the number of columns or the number of rows?
```{r,  eval=FALSE, tidy=TRUE}
## 1. Using RData files is the fastest -
## although you have to read the data in first.
## Set colClasses also produces an good speed-up.

##2. Setting colClasses R is no longer checking your data types.
## If your data is changing - for example it's coming from the web
## or a database, this may be problem.

##3. The results do depend on the number of columns, as this code demonstrates
N = c(1, 10, 100, 1000, 1000, 10000)
l = numeric(5)
for (i in seq_along(N)) {
  m = as.data.frame(matrix(runif(N[6]), ncol = N[i]))
  write.csv(m, file = "example.csv", row.names = FALSE)
  cc = rep("numeric", N[i])
  l[i] = system.time(
    read.csv("example.csv", colClasses = cc))[3]
}
l

## Notice that when we have a large number of columns,
## we get a slow down in reading in data set (even though
## we have specified the column classes). The reason for
## this slow down is that we are creating a data frame
## and each column has to be initialised with a particular class.
```




\newpage
## Practical 2


1. In this question, we'll compare matrices and data frames. Suppose we have a matrix, `d_m`
    ```{r, echo = TRUE}
    ##For fast computers
    #d_m = matrix(1:1000000, ncol=1000)
    ##Slower computers
    d_m = matrix(1:10000, ncol = 100)
    dim(d_m)
    ```
and a data frame `d_df`:
    ```{r, echo = TRUE}
    d_df = as.data.frame(d_m)
    colnames(d_df) = paste0("c", seq_along(d_df))
    ```
    + Using the following code, calculate the relative differences between selecting the first column/row of a data frame and matrix.
        ```{r,results='hide', tidy=FALSE, echo = TRUE}
        mark(d_m[1, ], d_df[1, ], d_m[, 1], d_df[, 1],
             relative = TRUE, 
             check = FALSE)
        ```
    Can you explain the result? Try varying the number of replications.
        ```{r,  tidy=TRUE}
        ## Two things are going on here
        ## 1. The very large difference when selecting columns and rows
        ## (in data frames)
        ## is because the data is stored in column major-order. Although the
        ## matrix is also stored in column major-order, because everything is
        ## the same type, we can efficiently select values.
        
        ##2. Matrices are also more memory efficient:
        m = matrix(runif(1e4), ncol = 1e4)
        d = data.frame(m)
        object.size(m)
        object.size(d)
        ```
      + When selecting columns in a data frame, there are a few different methods. For example,
        ```{r,results='hide', echo = TRUE}
        d_df$c10
        d_df[, 10]
        d_df[, "c10"]
        d_df[, colnames(d_df) == "c10"]
        ```
      Compare these four methods.
        ```{r, eval=FALSE}
        mark(d_df$c10, 
             d_df[, 10], 
             d_df[, "c10"],
             d_df[, colnames(d_df) == "c10"],
             iterations = 10000, 
             relative = TRUE)
        
        m = matrix(1:100000000, ncol = 10000)
        dim(m)
        
        mark(m[, 1], m[1, ],
             iterations = 10000, 
             relative = TRUE)
        ```
        ```{r,echo=FALSE}
        n = 1
        ```
2. Consider the following piece of code:
    ```{r,tidy=FALSE, echo = TRUE}
    a = NULL
    for (i in 1:n)
      a = c(a, 2 * pi * sin(i))
    ```
This code calculates the values:
\[
2\pi \sin(1), 2 \pi \sin(2), 2 \pi sin(3), \ldots, 2 \pi sin(n)
\]
and stores them in a vector. Two obvious ways of speeding up this code are:
    - Pre-allocate the vector `a` for storing your results.
    - Remove $2 \times \pi$ from the loop, i.e. at the end of the loop have the statement: `2*pi*a`.
Try the above techniques for speeding up the loop. Vary $n$ and plot your results.
3. R is an interpreted language; this means that the interpreter executes the program source code directly, statement by statement. Therefore, every function call takes time.^[This example is for illustrative proposes. Please don't start worrying about comments and brackets.] Consider these three examples:
    ```{r,cache=TRUE, tidy=FALSE, echo = TRUE}
    n = 1e6
    ## Example 1
    I = 0
    for (i in 1:n) {
      10
      I = I + 1
    }
    ## Example 2
    I = 0
    for (i in 1:n) {
      ((((((((((10))))))))))
      I = I + 1
    }
    ## Example 3
    I = 0
    for (i in 1:n) {
      ##This is a comment
      ##But it is still parsed
      ##So takes time
      ##But not a lot
      ##So don't worry!
      10
      I = I + 1
    }
    ```
Using the `mark()` function, time these three examples.



\newpage
## Practical 3: parallel programming


1.  To begin, load the `parallel` package and determine how many cores you have
    ```{r,results='hide', echo = TRUE}
    library("parallel")
    detectCores()
    ```
2. Run the parallel `apply` example in the notes.
  - On your machine, what value of `N` do you need to use to make the parallel code run quicker than the standard serial version?
  - When I ran the benchmarks, I didn't include the `makeCluster` and `stopCluster` functions calls. Include these calls in your timings. How does this affect your benchmarks?

3. Run the dice game Monte-Carlo example in the notes. Vary the parameter `M`.^[Try setting `M=50` and varying `N`.]


## Solutions

Solutions are contained within this package:
```{r,eval=FALSE, echo = TRUE}
library("jrEfficient")
vignette("solutions1", package = "jrEfficient")
```
